### 2021-04-14 paper.Rmd
new cluster image
tweaks to notation section

---
author:
  - name: Stuart Lee
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Department of Econometrics and Business Statistics, Monash University
    email: stuart.a.lee@monash.edu
  - name: Dianne Cook
    address: |
      | Department of Econometrics and Business Statistics, Monash University
    affiliation: Monash University
    email: dicook@monash.edu
  - name: Natalia da Silva
    address: |
      | Instituto de Estadística (IESTA), Universidad de la República
    affiliation: Universidad de la República
    email: natalia@iesta.edu.uy
  - name: Ursula Laa 
    affiliation: University of Natural Resources and Life Sciences
    address:  |
      | Institute of Statistics, University of Natural Resources and Life Sciences
    email:  ursula.laa@boku.ac.at
  - name: Nicholas Spyrison
    address: |
      | Faculty of Information and Technology, Monash University
    affiliation: Monash University
    email: nicholas.spyrison@monash.edu
  - name: Earo Wang
    address: |
      | Department of Statistics, The University of Auckland
    affiliation: The University of Auckland
    email: earo.wang@auckland.ac.nz
  - name: H. Sherry Zhang
    address: |
      | Department of Econometrics and Business Statistics, Monash University
    affiliation: Monash University
    email: huize.zhang@monash.edu
title:
  formatted: "A Review of the State-of-the-Art on Tours for Dynamic Visualisation of High-dimensional Data"
abstract: > 
  This article discusses a high-dimensional visualization technique called the tour, which can be used to view data in more than three dimensions. We review the theory and history behind the technique, as well as modern software developments and applications of the tour that are being found across the sciences and machine learning. 
keywords:
  formatted: ["tours", "data visualization", "high-dimensional data", "data science", "exploratory data analysis"]
output: 
  bookdown::pdf_document2:
    template: template/wires_template_rmd.tex
    keep_tex: true
    toc: false
bibliography: references.bib
biblio-style: apalike
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.show = 'hold',
  dpi = 300 #,
  # out.width = "\\textwidth",
  # fig.path = here::here('img/')
)
```

@@

\begin{figure*}[!t]
         \includegraphics[width=0.371\textwidth]{figures/cluster_example_r.pdf}
     \vfill
   \begin{minipage}{\linewidth}     
   \centering
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour1.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour2.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour3.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour4.png}
   \end{minipage}%
\caption{Comparison of structure perception between nonlinear dimension reduction from 10-d, using t-SNE (A) and a tour (B). There are six clusters, as seen in the t-SNE view, but the relative distance between the clusters is extremely varied. This can be seen in sample of tour projections shown. The two green clusters are (almost) spherical in shape, and very distance from the brown clusters. Three of the brown clusters are very close to each other (just visible in B.1), and all brown clusters are elliptical. The tour provides a more accurate rendering of the clusters in the high-dimensional space, and complements what is learned from the dimension reduction.}
\label{fig:tsne-tour}
\end{figure*}

@@

@@

A sequence of projections can be generated via different tour paths, i.e. grand tour, guided tour, manual tour, little tour, local tour, and most recently, slice tour. The focus of this section will be on the grand and guided tour, with manual and slice tour being introduced in the later sections of the review.

Grand tour, coined by @Asimov1985-xr, randomly samples target bases in the space and constructs an interpolation between them to produce a sequence of projections as in Figure  \ref{fig:notation-target-gen}. This tour path is suitable for an initial view of the data, however, as the dimension increases, it could take a long time before the grand tour finds a projection.

Instead of randomly generating targets, guided tour [@Cook1995-ae] finds its targets with a technique called projection pursuit [@Friedman1974-ck].  In projection pursuit, an index function is defined that is a subjective mapping from the projection $Y$ in $\mathbb{R}^{n \times d}$ to a scalar in $\mathbb{R}$. The guided tour iteratively optimizes the index function in the projection pursuit and produces the most interesting projection according to the index measure. The right panel of Figure \ref{fig:notation-target-gen} shows an example of a guided tour path.

@@

The smooth animation between the selected target frames is generated via a step-wise rotation in the high-dimensional space. There are two different methods to derive this rotation, first described in @Buja2005-cx., When interpolating between frames we rotate the starting basis onto the exact basis representation selected as a basis, and @Buja2005-cx describes three different methods to achieve this (based on decompositions of orthogonal matrices, Givens decompositions, or Householder decompositions) <!--SL : this sentence is confusing -->. Alternatively, when interpolating between planes the rotation is constructed towards any basis that represents the plane corresponding to the target frame, such that it follows the shortest path, i.e. the geodesic, avoiding any in-plane rotation. The geodesic interpolation is derived using singular value decomposition and is implemented in the `tourr` package.

<!-- SL: The above paragraph is very confusing to me, what is a plane and frame? what is a geodesic? why should it be a geodesic? -->

Typically the geodesic interpolation is preferred, since within-plane rotation is distracting and does not provide new information. An exception might be the case of a projection pursuit guided tour with an index function that is sensitive to these rotations [@Laa2020-zj].

For any method we need to ensure that each projection along the interpolated path is still orthonormal. As explained above this means a tour path is a trace along a Stiefel manifold. <!-- I still don't know if this adds anything, without it being defined --> In the case of one dimensional projections, the set of bases traces a path along the surface of a hypersphere of radius one, as illustrated in Figure \ref{fig:notation-target-gen}, and for two dimensional projections projections fall onto a torus shape. We can use a tour to visualize such a path, as shown in Figure \ref{fig:interpolation-2d}.

@@

Interactive systems like GGobi provide the option of adding sectioning information via linked brushing. Sectioning means we select points that fall in a section of the full parameter space, for example for the purpose of highlighting them in a tour display. This often reveals complementary information, and combining sections and projections can for example provide insights into the dimensionality of a data structure [@Furnas1994-dz].

@@

This review highlights the history of the grand tour and its modern extensions for the visualization of high dimensional numerical data. A rich ecosystem of tour methods are being actively developed and are finding applications in a broad range of areas such as machine learning and physics. Tours are being combined with interactive graphics to assist with cluster analysis (the spinifex and liminal packages), displays of projected data are augmented to allow the identification of local structures (the sage algorithm, slicing and section pursuit), and  new visualizations based on tour methods are used to better diagnose optimization procedures for projection pursuit (ferrn).

There are many possible future directions for tour research, one big limitation
of the current software ecosystem is that analysts are no longer able to perform interactions that were possible in software packages from the 80s and 90s. An interactive graphics system that integrated the tour methodology directly would be a big step forward. Such an interface could also be combined with methods for conditional model visualization, for example as implemented in the condviz package [@JSSv081i05], allowing the visual exploration of multivariate fitted models.

>>>>

---
author:
  - name: Stuart Lee
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Department of Econometrics and Business Statistics, Monash University
    email: stuart.a.lee@monash.edu
  - name: Dianne Cook
    address: |
      | Department of Econometrics and Business Statistics, Monash University
    affiliation: Monash University
    email: dicook@monash.edu
  - name: Natalia da Silva
    address: |
      | Instituto de Estadística (IESTA), Universidad de la República
    affiliation: Universidad de la República
    email: natalia@iesta.edu.uy
  - name: Ursula Laa 
    affiliation: University of Natural Resources and Life Sciences
    address:  |
      | Institute of Statistics, University of Natural Resources and Life Sciences
    email:  ursula.laa@boku.ac.at
  - name: Nicholas Spyrison
    address: |
      | Faculty of Information and Technology, Monash University
    affiliation: Monash University
    email: nicholas.spyrison@monash.edu
  - name: Earo Wang
    address: |
      | Department of Statistics, The University of Auckland
    affiliation: The University of Auckland
    email: earo.wang@auckland.ac.nz
  - name: H. Sherry Zhang
    address: |
      | Department of Econometrics and Business Statistics, Monash University
    affiliation: Monash University
    email: huize.zhang@monash.edu
title:
  formatted: "A Review of the State-of-the-Art on Tours for Dynamic Visualisation of High-dimensional Data"
abstract: > 
  This article discusses a high-dimensional visualization technique called the tour, which can be used to view data in more than three dimensions. We review the theory and history behind the technique, as well as modern software developments and applications of the tour that are being found across the sciences and machine learning. 
keywords:
  formatted: ["tours", "data visualization", "high-dimensional data", "data science", "exploratory data analysis"]
output: 
  bookdown::pdf_document2:
    template: template/wires_template_rmd.tex
    keep_tex: true
    toc: false
bibliography: references.bib
biblio-style: apalike
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.show = 'hold',
  dpi = 300,
  cache = FALSE
  # out.width = "\\textwidth",
  # fig.path = here::here('img/')
)
```

@@

\begin{figure*}[!t]
         \includegraphics[width=0.371\textwidth]{figures/cluster_example.pdf}
     \vfill
   \begin{minipage}{\linewidth}     
   \centering
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour1.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour2.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour3.png}
         \includegraphics[width=0.24\textwidth]{figures/cluster_example_new_tour4.png}
   \end{minipage}%
\caption{Comparison of structure perception between nonlinear dimension reduction from 10-d, using t-SNE (A) and a tour (B). There are six clusters, as seen in the t-SNE view, but the relative distance between the clusters is extremely varied. This can be seen in sample of tour projections shown. The two green clusters are (almost) spherical in shape, and very distance from the brown clusters. Three of the brown clusters are very close to each other (just visible in B.1), and all brown clusters are elliptical. The tour provides a more accurate rendering of the clusters in the high-dimensional space, and complements what is learned from the dimension reduction.}
\label{fig:tsne-tour}
\end{figure*}

@@

\begin{figure*}
   \centering
         \includegraphics[width=\textwidth]{figures/tour_path.png}
\caption{Illustration of a tour path, indicating target planes, and interpolation planes. (Adapted from Buja et al. 2005, Figure 1.) }
\label{fig:tour-path}
\end{figure*}

@@

A tour path, that is a sequence of projections, can be generated by geodesic interpolation between a set of target planes (Figure \ref{fig:tour-path}). Different methods for choosing target planes provide different tour paths. The grand tour is generated using randomly selected target. A guided tour is generated by choosing particularly structured projection from a projection pursuit optimisation. The little tour uses all variable bases as targets, and a local tour rocks back and forth from a particular plane to randomly chosen targets in a small neighbourhood.

Figure \ref{fig:notation-target-gen} shows a representation of 1D tour paths of 5D data, drawn on a PCA space. A grand tour path is on the left and a guided tour path is on the right. The grand tour path should be more wide-spread because it is attempting to show all possible projections, if left to run long enough. The guided tour path should be short, as the projection pursuit optimisation zeroes on the optimally interesting projection.

There are other tour types which don't fit this style of tour. The manual tour [@cook_manual_1997] allows the user to change the projection coefficients manually, to rotate a variable into and out of a project, and is discussed later. The recently developed slice tour [@Laa2020-js] can be applied to any of the above tour types. It displays a slice through the orthogonal space, as opposed to a projection, and is explained in more details later.

@@

The smooth progression in the tour path is due to geodesic interpolation between the target planes. This takes into consideration two important aspects: (1) maintains the orthonormality of the projection bases, (2) contains all the rotation to be between planes, not a particular basis in any plane. The first is clearly important because it ensures that we a looking at low-dimensional projection of the data always. The latter is harder to explain, but really important from a visual perspective. It stops any within-plane spin, and could be considered stabilising the view. More details explanation of this can be found in @Buja2005-cx.

Figure \ref{fig:interpolation-2d} shows tour paths of 2D projections of 6D data. The space of all tour paths is a high-dimensional torus, as represented by the grey points. The tour paths are shown in green and brown, and each dot indicates a projection in the sequence.  The plots in this figure could be considered to be a tour looking at itself, because each plot is a selected view from a tour of the torus with the paths overlaid.

@@

Interactive systems like GGobi [@Swayne2003] provide the option of adding sectioning information via linked brushing. Sectioning means we select points that fall in a section of the full parameter space, for example for the purpose of highlighting them in a tour display. This often reveals complementary information, and combining sections and projections can for example provide insights into the dimensionality of a data structure [@Furnas1994-dz].

@@

This review highlights the history of tour methods, and modern extensions for the visualization of high dimensional numerical data. The rich ecosystem of tours is being actively developed and applied to a broad range of problems including machine learning. When tours are be combined with some interactivity it can assist with analyses like clustering (the spinifex and liminal packages).  Modified displays of projected data can be used to expose local structures (the sage algorithm, slicing and section pursuit). New algorithm data collection inside the tour code is being used to better diagnose optimization procedures for projection pursuit (ferrn).

There are many possible future directions for tour research.  One big challenge for software engineering is to seamlessly embed the tour with interactivity. Currently, there is no implementation that analysts can use to perform tasks that were easy in software available in the 80s and 90s. More contributions in sectioning as an alternative to projection, could be broadly useful. An example is conditional model visualization, as implemented in the condviz package [@JSSv081i05], which assists with visual exploration of multivariate fitted models. It could be interesting to define tour paths for different types of nonlinear subspaces, or non-Euclidean space.
